
##Example code for analysis performed on November 
```{r}
November2015 <- read.csv("November2015.csv")
#Adding column names to the data
colnames(November2015)<-c("Vehicle_ID","Time", "Latitude","Longitude", "Speed", "Direction", "OffRouteStatus", "CommStatus", "OperationalStatus", "Server_Time", "Route","Trip", "Inbound_Outbound", "Deviation", "Onboard", "Vehicle_Name", "Run_Id", "Run_Name", "Stop_Name", "Operator_Record_Id", "Route_Name", "Stop_Report", "Scheduled_Headway", "Target_Headway", "Alarm_State", "GPSStatus", "Boards", "Alights", "Confidence_Level", "Message_Type_Id", "Stop_Dwell_Time", "PTV_Health_Alert", "Stop_Id", "StationaryStatus", "StationaryDuration", "VehicleStatusID", "Veh_Type_Id", "Block_Farebox_Id", "OdometerValue", "MDTFlags")
```

##Creating rush hours as well as weekend/weekday identifier variables
```{r}
#Breaking up the time variable and adding it to the subset
November2015$year<-as.numeric(substring(November2015$Time,1,4))
November2015$month<-as.numeric(substring(November2015$Time,6,7))
November2015$day<-as.numeric(substring(November2015$Time,9,10))
November2015$hour<-as.numeric(substring(November2015$Time,12,13))
November2015$minutes<-as.numeric(substring(November2015$Time,15,16))
seconds<-substring(November2015$Time,18,23)
seconds<-as.numeric(gsub("\\.", "", seconds))
seconds<-signif(seconds,digits=2)
November2015$seconds<-as.numeric(substring(seconds,1,(nchar(seconds))-3))

November2015$dayofweek<-weekdays(as.Date(November2015$Time))
November2015$dayofweek<-as.factor(November2015$dayofweek)

November2015$weekcat[November2015$dayofweek == "Friday"] <- "weekend"
November2015$weekcat[November2015$dayofweek == "Saturday"] <- "weekend"
November2015$weekcat[November2015$dayofweek == "Sunday"] <- "weekend"
November2015$weekcat[November2015$dayofweek == "Thursday"] <- "weekday"
November2015$weekcat[November2015$dayofweek == "Wednesday"] <- "weekday"
November2015$weekcat[November2015$dayofweek == "Tuesday"] <- "weekday"
November2015$weekcat[November2015$dayofweek == "Monday"] <- "weekday"

#In the dataset November2015, make the changes to add rush hours
November2015$rush[November2015$hour == 16|November2015$hour == 17|November2015$hour == 18] <- "AfternoonRush"
November2015$rush[November2015$hour == 7|November2015$hour == 8|November2015$hour == 9] <- "MorningRush"
November2015$rush[November2015$hour == 10|November2015$hour == 11|November2015$hour == 12] <- "AfternoonNonRush"
November2015$rush[November2015$hour == 20|November2015$hour == 21|November2015$hour == 22] <- "EveningNonRush"

LateNovember33 <- subset(November2015, OperationalStatus == 1 & Route_Name == 33) #subset all late 31 buses
LateNovember33$weekcat<-as.factor(LateNovember33$weekcat)

#How much so rush hours deviate by from non rush hours in weekday vs weekend?
mean(~Deviation, data=NovemberAfternoon) #deviation for afternoon rush hours (1.955)
mean(~Deviation, data=NovemberMorning) #deviation for morning rush hours (1.244)
mean(~Deviation, data=NovemberNonRush1) #deviation is 1.515328
mean(~Deviation, data=NovemberNonRush2) #deviation is 1.578127

LateNovemberDeviWeek <- subset(LateNovember, weekcat == "weekday")
LateNovemberDeviWeekend <- subset(LateNovember, weekcat == "weekend")

LateNovember33 %>%
  group_by(weekcat, rush) %>%
  summarise(mean(Deviation), sd(Deviation))

LateNovember33 %>%
  group_by(Route_Name) %>%
  summarise(mean(Deviation), sd(Deviation))

LateNovemberDeviWeekend %>%
  group_by(Route_Name) %>%
  summarise(mean(Deviation))

SampleObs <- read.csv("SampleObs.csv")
nrow(SampleObs)
View(SampleObs)
```

##How does average speed, lateness change with different years
```{r}
#returns average speed of the data inputted
speed<-function(data) {
  speeddata<-data[5]
  return (mean(speeddata)) #max speed for January2015 is 66
}

#returns number of late 
Late<-function(data){
  ((tally(~OperationalStatus)[1]+tally(~OperationalStatus)[2])/(tally(~OperationalStatus)[1]+tally(~OperationalStatus)[2]+tally(~OperationalStatus)[3]))
}

#returns percentage of ontime and early PVTA rides
Ontime <- function(data){
  return ((tally(~OperationalStatus)[1]+tally(~OperationalStatus)[2])/(tally(~OperationalStatus)[1]+tally(~OperationalStatus)[2]+tally(~OperationalStatus)[3]))*100
}

#returns bad communication (bad gps, inactive) inputting year
Comm<-function(data){
  return ((tally(~CommStatus)[1]+tally(~CommStatus)[2]+tally(~CommStatus)[4])/(tally(~CommStatus)[1]+tally(~CommStatus)[2]+tally(~CommStatus)[3]+tally(~CommStatus)[4]))*100
}
```

##Stepwise regression in both forward and backward directions 
```{r}
full <- lm(Speed ~ Alights+Boards+Onboard+OperationalStatus+CommStatus+GPSStatus+OffRouteStatus)
summary(full)
step(full, direction = "both")
```

##Analyze Deviation in PVTA routes
```{r}
#Analyze route deviation by routes (group_by operator)
November2015 %>%
  group_by(Route_Name) %>%
  summarise(DevRoute = mean(Deviation), numrow = n())

November45$GPSStatus <- as.factor(November45$GPSStatus)
November45<-subset(November2015, Route_Name == "45")
mod1 <- lm(Deviation ~ Onboard + Speed + CommStatus+ Alights + Boards + GPSStatus + OperationalStatus)
summary(mod1)
```


##Creating holiday status predictor
```{r}
LateNovember33$holiday<-ifelse(LateNovember33$day>24 &LateNovember33$day<31,"holiday","nonholiday")

#Look at holidays for March since no run for November
LateMarch33$holiday<-ifelse(LateMarch33$day>13 & LateMarch33$day<24,"holiday","nonholiday")
tally(holiday~day, data=LateMarch33)
```

##Visualizing bus deviation from schedule by rush hours 
```{r}

p1<-bwplot(Deviation ~ rush  |  weekcat, data=LateFeb33, main="Late Feb33 Deviations", type=c("p","r"))
p2<-bwplot(Deviation ~ rush  |  weekcat, data=LateOctober33, main="Late Oct33 Deviations", type=c("p","r"))

print(p1, position = c(0, 0, 0.5, 1), more = TRUE)
print(p2, position = c(0.5, 0, 1, 1))


p1<-bwplot(Deviation ~ rush  |  weekcat, data=LateMarch33, main="Late March33 Deviations", type=c("p","r"))
p2<-bwplot(Deviation ~ rush  |  weekcat, data=LateNovember33, main="Late Nov33 Deviations", type=c("p","r"))

print(p1, position = c(0, 0, 0.5, 1), more = TRUE)
print(p2, position = c(0.5, 0, 1, 1))
```

